---
title: "FST Distribution Plots"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(dplyr)
```

# 1. define a function to plot distribution of per-site minor af per population (FR)
```{r}
plot_fst_distribution <- function(chrom) {
  # Load data
  fst_file <- paste0("../data/output/afs_FR_", chrom, ".txt")
  
  fst_data <- read.table(fst_file, header = FALSE)
  
# Combine data into a single data frame
  fst_data_combined <- data.frame(FST = fst_data$V1, Version = "minor allele frequency")
  
  # Calculate top percentiles
  percentiles <- fst_data_combined %>%
    group_by(Version) %>%
    summarize(
      p90 = quantile(FST, 0.90),
      p95 = quantile(FST, 0.95)
    )
  
  # Plot
  p <- ggplot(fst_data_combined, aes(x = FST, fill = Version)) +
    geom_density(alpha = 0.5, bw = 0.01) +
    geom_vline(data = percentiles, aes(xintercept = p90, color = Version, linetype = "90th Percentile"), size = 1) +
    geom_vline(data = percentiles, aes(xintercept = p95, color = Version, linetype = "95th Percentile"), size = 1) +
    scale_linetype_manual(name = "Percentiles", values = c("90th Percentile" = "dashed", "95th Percentile" = "dotted")) +
    labs(title = paste("MAF Distribution for", chrom), x = "MAF", y = "Density") +
    theme_minimal() +
    scale_x_continuous(breaks = scales::pretty_breaks(n = 10), expand = c(0, 0)) +
    scale_y_continuous(breaks = scales::pretty_breaks(n = 10), expand = c(0, 0)) +
    theme(
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      axis.ticks = element_line(size = 0.75),       # Add axis ticks
      axis.ticks.length = unit(0.25, "cm"),         # Adjust tick length
      axis.line = element_line(size = 0.75, color = "black"), # Add axis lines
      panel.background = element_rect(fill = "white", color = NA), # White background
      plot.background = element_rect(fill = "white", color = NA)   # White background
    )
  # Save the plot
  ggsave(paste0("../data/output/MAF_FR_distribution_", chrom, "_empirical.png"), plot = p, width = 6, height = 4, dpi = 300)
  return(p)
}
```

# 1. define a function to plot distribution of per-site minor af per population (ZI)
```{r}
plot_fst_distribution <- function(chrom) {
  # Load data
  fst_file <- paste0("../data/output/afs_ZI_", chrom, ".txt")
  
  fst_data <- read.table(fst_file, header = FALSE)
  
# Combine data into a single data frame
  fst_data_combined <- data.frame(FST = fst_data$V1, Version = "minor allele frequency")
  
  # Calculate top percentiles
  percentiles <- fst_data_combined %>%
    group_by(Version) %>%
    summarize(
      p90 = quantile(FST, 0.90),
      p95 = quantile(FST, 0.95)
    )
  
  # Plot
  p <- ggplot(fst_data_combined, aes(x = FST, fill = Version)) +
    geom_density(alpha = 0.5, bw = 0.01) +
    geom_vline(data = percentiles, aes(xintercept = p90, color = Version, linetype = "90th Percentile"), size = 1) +
    geom_vline(data = percentiles, aes(xintercept = p95, color = Version, linetype = "95th Percentile"), size = 1) +
    scale_linetype_manual(name = "Percentiles", values = c("90th Percentile" = "dashed", "95th Percentile" = "dotted")) +
    labs(title = paste("MAF Distribution for", chrom), x = "MAF", y = "Density") +
    theme_minimal() +
    scale_x_continuous(breaks = scales::pretty_breaks(n = 10), expand = c(0, 0)) +
    scale_y_continuous(breaks = scales::pretty_breaks(n = 10), expand = c(0, 0)) +
    theme(
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      axis.ticks = element_line(size = 0.75),       # Add axis ticks
      axis.ticks.length = unit(0.25, "cm"),         # Adjust tick length
      axis.line = element_line(size = 0.75, color = "black"), # Add axis lines
      panel.background = element_rect(fill = "white", color = NA), # White background
      plot.background = element_rect(fill = "white", color = NA)   # White background
    )
  # Save the plot
  ggsave(paste0("../data/output/MAF_ZI_distribution_", chrom, "_empirical.png"), plot = p, width = 6, height = 4, dpi = 300)
  return(p)
}
```


# 1. define a function to plot distribution of per-site total heterozygosity across populations
```{r}
plot_fst_distribution <- function(chrom) {
  # Load data
  fst_file <- paste0("../data/output/heterozygosity_", chrom, ".txt")
  
  fst_data <- read.table(fst_file, header = FALSE)
  
# Combine data into a single data frame
  fst_data_combined <- data.frame(FST = fst_data$V1, Version = "SNP heterozygosity")
  
  # Calculate top percentiles
  percentiles <- fst_data_combined %>%
    group_by(Version) %>%
    summarize(
      p90 = quantile(FST, 0.90),
      p95 = quantile(FST, 0.95)
    )
  
  # Plot
  p <- ggplot(fst_data_combined, aes(x = FST, fill = Version)) +
    geom_density(alpha = 0.5, bw = 0.01) +
    geom_vline(data = percentiles, aes(xintercept = p90, color = Version, linetype = "90th Percentile"), size = 1) +
    geom_vline(data = percentiles, aes(xintercept = p95, color = Version, linetype = "95th Percentile"), size = 1) +
    scale_linetype_manual(name = "Percentiles", values = c("90th Percentile" = "dashed", "95th Percentile" = "dotted")) +
    labs(title = paste("Heterozygosity Distribution for", chrom), x = "Heterozygosity", y = "Density") +
    theme_minimal() +
    scale_x_continuous(breaks = scales::pretty_breaks(n = 10), expand = c(0, 0)) +
    scale_y_continuous(breaks = scales::pretty_breaks(n = 10), expand = c(0, 0)) +
    theme(
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      axis.ticks = element_line(size = 0.75),       # Add axis ticks
      axis.ticks.length = unit(0.25, "cm"),         # Adjust tick length
      axis.line = element_line(size = 0.75, color = "black"), # Add axis lines
      panel.background = element_rect(fill = "white", color = NA), # White background
      plot.background = element_rect(fill = "white", color = NA)   # White background
    )
  # Save the plot
  ggsave(paste0("../data/output/heterozygosity_distribution_", chrom, "_empirical.png"), plot = p, width = 6, height = 4, dpi = 300)
  return(p)
}
```


# 1. define a function to plot FST distribution
```{r}
plot_fst_distribution <- function(chrom) {
  # Load data
  fst_file <- paste0("../data/output/fst_", chrom, ".txt")
  
  fst_data <- read.table(fst_file, header = FALSE)
  
# Combine data into a single data frame
  fst_data_combined <- data.frame(FST = fst_data$V1, Version = "SNP FST")
  
  # Calculate top percentiles
  percentiles <- fst_data_combined %>%
    group_by(Version) %>%
    summarize(
      p90 = quantile(FST, 0.90),
      p95 = quantile(FST, 0.95)
    )
  
  # Plot
  p <- ggplot(fst_data_combined, aes(x = FST, fill = Version)) +
    geom_density(alpha = 0.5, bw = 0.01) +
    geom_vline(data = percentiles, aes(xintercept = p90, color = Version, linetype = "90th Percentile"), size = 1) +
    geom_vline(data = percentiles, aes(xintercept = p95, color = Version, linetype = "95th Percentile"), size = 1) +
    scale_linetype_manual(name = "Percentiles", values = c("90th Percentile" = "dashed", "95th Percentile" = "dotted")) +
    labs(title = paste("FST Distribution for", chrom), x = "FST", y = "Density") +
    theme_minimal() +
    scale_x_continuous(breaks = scales::pretty_breaks(n = 10), expand = c(0, 0)) +
    scale_y_continuous(breaks = scales::pretty_breaks(n = 10), expand = c(0, 0)) +
    theme(
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      axis.ticks = element_line(size = 0.75),       # Add axis ticks
      axis.ticks.length = unit(0.25, "cm"),         # Adjust tick length
      axis.line = element_line(size = 0.75, color = "black"), # Add axis lines
      panel.background = element_rect(fill = "white", color = NA), # White background
      plot.background = element_rect(fill = "white", color = NA)   # White background
    )
  # Save the plot
  ggsave(paste0("../data/output/fst_distribution_", chrom, "_empirical.png"), plot = p, width = 6, height = 4, dpi = 300)
  return(p)
}
```

# 1. define a function to plot FST distribution (biallelic)
```{r}
plot_fst_distribution <- function(chrom) {
  # Load data
  fst_file <- paste0("../data/output/fst_", chrom, "_biallelic.txt")
  
  fst_data <- read.table(fst_file, header = FALSE)
  
# Combine data into a single data frame
  fst_data_combined <- data.frame(FST = fst_data$V1, Version = "SNP FST")
  
  # Calculate top percentiles
  percentiles <- fst_data_combined %>%
    group_by(Version) %>%
    summarize(
      p90 = quantile(FST, 0.90),
      p95 = quantile(FST, 0.95)
    )
  
  # Plot
  p <- ggplot(fst_data_combined, aes(x = FST, fill = Version)) +
    geom_density(alpha = 0.5, bw = 0.01) +
    geom_vline(data = percentiles, aes(xintercept = p90, color = Version, linetype = "90th Percentile"), size = 1) +
    geom_vline(data = percentiles, aes(xintercept = p95, color = Version, linetype = "95th Percentile"), size = 1) +
    scale_linetype_manual(name = "Percentiles", values = c("90th Percentile" = "dashed", "95th Percentile" = "dotted")) +
    labs(title = paste("FST Distribution for", chrom), x = "FST", y = "Density") +
    theme_minimal() +
    scale_x_continuous(breaks = scales::pretty_breaks(n = 10), expand = c(0, 0)) +
    scale_y_continuous(breaks = scales::pretty_breaks(n = 10), expand = c(0, 0)) +
    theme(
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      axis.ticks = element_line(size = 0.75),       # Add axis ticks
      axis.ticks.length = unit(0.25, "cm"),         # Adjust tick length
      axis.line = element_line(size = 0.75, color = "black"), # Add axis lines
      panel.background = element_rect(fill = "white", color = NA), # White background
      plot.background = element_rect(fill = "white", color = NA)   # White background
    )
  # Save the plot
  ggsave(paste0("../data/output/fst_distribution_", chrom, "_empirical_biallelic.png"), plot = p, width = 6, height = 4, dpi = 300)
  return(p)
}
```

# 2. plot FST distribution
```{r}
for (chr in c("chrX", "chr2L", "chr2R", "chr3L", "chr3R", "combined")) {
  plot_fst_distribution(chr)
}
```